Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
massage the comparison section

\newcommand\naive{na\"\i ve}
\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\usepackage{graphicx}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tool or library it uses is \texttt{zlib}.
Its initial development took \emph{two years}.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++.
Its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell
Its initial development took about three months.
\remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}

Matt uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
It relies on no external libraries beyond \texttt{zlib},
but contains 1,000 lines of protein-structure code that ought
to be in a library.

We~modified Matt to use information about sequences as well as structure.
We~thought the modification would take three months, 
but it took most of a~year.
This modification added 2,000 lines of code, and relies on
system calls to external programs (sequence aligners) that we did not write.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.
We~had hoped to add support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

Like many researchers, we~and our colleagues often write code hastily.
Hastily written C++ code can get a job done, 
but we have found it expensive in the long run.
Our most painful experience was with an implementation of
\textit{multidimensional}
dynamic programming which was hastily added to the SMURF homology detector

The implementation took 9,000~lines, including a single function of
over 700 lines.
When we added simulated evolution to this code, a relatively minor
enhancement \cite{Daniels:2012}, 
just understanding the existing code took several months.
Part of SMURF's code includes the 1,000 lines of protein-structure code
copied from the previous Matt software.

For comparison, 
building the first functioning version of MRFy took three months of part-time work.
And we expect that higher-order functions will make extending MRFy
easy.
MRFy currently contains 1,200 lines of Haskell code, though it relies heavily on
a number of libraries beyond the standard Prelude.
We found that adapting our software to explore a new search heuristic was typically
less than a day's work.
There are two major research ideas we wish to add to MRFy eventually, one of which
\remark{how much should I say about profiles and alpha helices? ---NMD}
involves looking at another type of nonlocal dependency (\textit{alpha helices} in contrast
to the current beta strands).
We expect that thanks to algebraic data types and typeclasses, these
enhancements will be at 
most a few weeks of work; we do not believe these enhancements would
even be feasible in SMURF. 
Indeed, Daniels tried to add another of these enhancements to SMURF
two years ago and gave up due 
to the difficulty of the implementation.
\remark{The above needs a lot of revision. It is a shitty first draft of a /P/. ---NMD}

@@

When we look beyond our own group to computational biology more
broadly, our experiences with other software is better.
Little of it is written in functional languages, 
but much of the software shared by the community is excellent.
MRFy's training component was derived from that of~HMMER,
and
working with HMMER 
codebase was pleasant;
data structures and their
mutators are well documented. 
There is a 
BioHaskell library, part of which we use,
but it is not nearly as 
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that tools for computational biology in
Haskell continue to mature.

>>>>

\newcommand\naive{na\"\i ve}
\usepackage{url}
\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\usepackage{graphicx}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

When performance has mattered, the members of our group, like other
computational biologists, have used~C++.
We~compare our experience with three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tool or library it uses is \texttt{zlib}.
Its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries.
Its initial development took a year and a half.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands.
It~comprises about 1,200 lines of~Haskell
It~uses several external tools and libraries, of which the most
notable are PADS-Haskell, the \path{Data.MemoCombinators} library,
the BioHaskell bioinformatics library, and the \path{Data.Vector} library.
Its initial development took about three months.
\remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

We~modified Matt to use information about sequences as well as structure.
The modification added 2,000 lines of code, and it calls
external sequence aligners that we did not write.
We~thought the modification would take three months, 
but it took most of a~year.
Matt uses such
data structures such as mutable oct-trees, vectors, and arrays.
It~uses clever pointer arithmetic.
The mutable data structures were difficult to 
repurpose, and the pointer arithmetic was too clever for its own good: 
nearly every change resulted in new segfaults.

We~had hoped to extend Matt further, with support for partial alignments,
which we~expected to require only 
a cosmetic manipulation of the 
output, but
this feature wound up
 requiring deep information about some data structures,
and we had to give~up.
We~believe we could write an equivalent tool in~Haskell,
with most of Matt's performance, in at most nine months.

Our most painful experience was adding ``simulated evolution''
to~SMURF \cite{Daniels:2012}. 
Although simulated evolution represents a relatively minor
enhancement,
just understanding the existing code took several months.

We~built MRFy quickly, and we expect that
higher-order functions will MRFy easy to extend.
Each new addition to MRFy's stochastic search has taken at most a day
to implement.

%
% An Experience Report is about what you actually did, not what you
% wish to do :-)
%
%%  There are two major research ideas we wish to add to MRFy eventually, one of which
%%  \remark{how much should I say about profiles and alpha helices? ---NMD}
%%  involves looking at another type of nonlocal dependency (\textit{alpha
%%  helices} in contrast 
%%  to the current beta strands).
%%  We expect that thanks to algebraic data types and typeclasses, these
%%  enhancements will be at 
%%  most a few weeks of work; we do not believe these enhancements would
%%  even be feasible in SMURF. 
%%  Indeed, Daniels tried to add another of these enhancements to SMURF
%%  two years ago and gave up due 
%%  to the difficulty of the implementation.
%%  \remark{The above needs a lot of revision. It is a shitty first
%%  draft of a /P/. ---NMD}

@@

Looking beyond our research group to computational biology more
broadly, our experience with other software is better.
Little of it is written in functional languages, 
but much of the software shared by the community is excellent.
MRFy's training component was derived from that of~HMMER,
and
working with HMMER 
codebase was pleasant;
data structures and their
mutators are well documented. 
There is a 
BioHaskell library, part of which we use,
but it is not nearly as 
complete as BioPython or BioRuby, which are heavily used in the community.
We hope that tools for computational biology in
Haskell continue to mature.
